Distributional Trees for Circular Data

### 2019-02-05 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
some further streamlining
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2436 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 92af93b3c57911048ea93b0d2293237ee0d34f5a

% Brief abstract of the paper:
\abstract{
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates along with interactions automatically. For illustration, hourly wind 
direction forecasts are obtained at Innsbruck Airport based on a set of meteorological
measurements.
}

@@

In each subgroup maximum likelihood estimators $\hat \mu$ and $\hat \kappa$ are obtained
by maximizing the corresponding log-likelihood $\ell(\mu, \kappa; y) = \log(f_\mathrm{vM}(y;\mu, \kappa))$. 
The model scores are calculated by  $s(y; \mu, \kappa) = (\partial_{\mu} \ell(\mu, \kappa; y),
\partial_{\kappa} \ell(\mu, \kappa; y))$. In a subgroup of size $n$, evaluating the scores
at the individual observations and parameter estimates $s(y_i; \hat{\mu}, \hat{\kappa})$
yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
respectively.
Hence MOB can test for parameter instabilities in the model by assessing whether the scores
change along with the available covariates. If so, by maximizing a partitioned likelihood
the parameter instabilities are incorporated into the model. This procedure is repeated recursively
until there are no significant parameter instabilities or until another stopping criterion
is met (e.g., concerning subgroup sizes or depth of the tree).

@@

\begin{figure}[!ht]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
\caption{Fitted distributional tree for a circular response employing the von
Mises distribution. For each terminal node, the empirical density and the 
fitted density distribution is displayed in gray and red, respectively. The
estimated location parameter is depicted as red hand and conveys the direction from
which the wind is blowing. The chosen split variables are wind direction
(dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$), and pressure gradients
along the upper (pdiff\_up; $\text{hPa}$) and lower
(pdiff\_low; $\text{hPa}$) Inn Valley where positive values point in the
direction of Innsbruck.}
\label{fig:tree} \end{figure}

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision making processes and risk management,
e.g., in air traffic management or renewable energy production. This study
presents an application of circular regression trees  on hourly wind direction
forecasts at Innsbruck Airport. Innsbruck lies at the bottom of the Inn
Valley, a deep mountain valley in the Eastern European Alps orientated parallel
to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered
in the North with an approx. 2300\,m high mountain ridge and opens to the South
into the crest-perpendicular Wipp Valley. As potential split variables we
employ several 1-hourly lagged observed weather variables, amongst others, the
pressure gradient along the upper and lower Inn Valley, and the wind direction and
wind speed at Innsbruck Airport.

As an illustration of the tree-based wind direction forecasts at Innsbruck
Airport, a distributional tree of depth three is depicted in Figure~\ref{fig:tree}
showing the circular histogram and fitted response distribution for each terminal node.
According to the fitted location parameter the following wind regimes can be summarized:
Firstly, up-valley winds blowing from the valley mouth towards the upper valley
(from east to west, nodes 4 and 5). Secondly, downslope winds blowing across
the Alpine crest along the Wipp Valley towards Innsbruck (from south-east to
north-west, nodes 7 and 8). Thirdly, down-valley winds blowing in the direction
of the valley mouth (from west to east, nodes 12, 14, and 15). The terminal node
11 captures the cases with rather low wind speeds which typically do not show a clear 
wind direction or distinct wind regime as implied by a small concentration 
parameter.
In general, winds typically blow from areas of high
pressure to areas of low pressure where strong pressure gradients yield high
wind speeds and a rather well-defined wind direction. Thus, the splits in the
lagged wind direction perform a clustering into the different wind regimes
identified by the fitted location parameter, and the additional splits in
pressure gradients or wind speed divide the predictions according to the
expected consistency of the wind direction expressed by the fitted
concentration parameter. This can be seen in the terminal nodes with strong
pressure gradients or high wind speeds where most training cases are classified
correctly and therefore the concentration parameter as an indicator of the
certainty of the forecast is rather large.

@@

This study shows that applying distributional trees on circular data provides
easily interpretable models clearly illustrating the impact of given covariates
and their possible interactions on the distribution of the dependent variable.
Moreover, covariates are selected automatically such that no prespecification
of the model is required which can be a challenging task, particularly for a
meteorological quantity such as wind direction with high variation and many
possible influencing factors. Additionally, the employed distributional trees
allow for abrupt changes in the response variable, such as e.g., a wind shift due to
a turn of the the pressure gradient within the Inn Valley.

A possible extension of the method as presented is to build an ensemble of
distributional trees to regularize and stabilize the model (see, e.g.,
distributional forests, Schlosser et.~al, 2018). 
Furthermore, in order to properly account not only for the circular properties of 
the response but also of the regressors, extensions in the search of split points 
would be feasible. For example, searching for two split points defining a split 
interval on the circular scale instead of a single split point on the linear scale 
could reduce the complexity of the model and possibly improve the results.

\medskip

\textit{Computational Details:} The employed \textsf{R} packages are available on \textsf{R}-Forge at
https://R-Forge.R-project.org/projects/partykit/.

@@

\item[Jammalamadaka, S. R., and Sengupta, A.] (2001).
     Topics in Circular Statistics.
     {\it World Scientific}, {\bf 5}.

@@

\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
     {\it arXiv.org E-Print Archive}, https://arxiv.org/abs/1804.02921v2.

>>>>

% Brief abstract of the paper:
\abstract{
For probabilistic modeling of circular data the von Mises distribution is widely used.
To capture how its parameters change with covariates, a regression tree model
is proposed as an alternative to more commonly-used additive models. The resulting
distributional trees are easy to interpret, can detect non-additive effects, and select
covariates and their interactions automatically. For illustration, hourly wind 
direction forecasts are obtained at Innsbruck Airport based on a set of meteorological
measurements.
}

@@

In each subgroup maximum likelihood estimators $\hat \mu$ and $\hat \kappa$ are obtained
by maximizing the corresponding log-likelihood $\ell(\mu, \kappa; y) = \log(f_\mathrm{vM}(y;\mu, \kappa))$. 
The model scores are given by  $s(y; \mu, \kappa) = (\partial_{\mu} \ell(\mu, \kappa; y),
\partial_{\kappa} \ell(\mu, \kappa; y))$. In a subgroup of size $n$, evaluating the scores
at the individual observations and parameter estimates $s(y_i; \hat{\mu}, \hat{\kappa})$
yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
respectively.
Hence MOB or CTree can assess whether the scores change along with the available covariates.
If so, by maximizing a partitioned likelihood
the parameter instabilities are incorporated into the model. This procedure is repeated recursively
until there are no significant parameter instabilities or until another stopping criterion
is met (e.g., subgroup size or tree depth).

@@

\begin{figure}[!ht]\centering
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates employed are
wind direction (dd; $\text{rad}$), wind speed (ff; $\text{ms}^{-1}$),
and pressure gradients along the upper (pdiff\_up; $\text{hPa}$) and lower
(pdiff\_low; $\text{hPa}$) Inn Valley.}
\label{fig:tree} \end{figure}

Wind is a classical circular quantity and accurate forecasts of wind direction
are of great importance for decision-making processes and risk management,
e.g., in air traffic management or renewable energy production. This study
employs circular regression trees to obtain hourly wind direction
forecasts at Innsbruck Airport. Innsbruck lies at the bottom of the Inn
Valley, a deep mountain valley in the Eastern European Alps orientated parallel
to the Alpine crest. At Innsbruck, the west-east aligned Inn Valley is bordered
in the North by an approx.~2300\,m high mountain ridge and opens to the South
into the crest-perpendicular Wipp Valley. Hence, pressure gradients along the
upper and lower Inn Valley are considered as covariates along with other
meteorological measurements (lagged by one hour) such as wind direction and
wind speed at Innsbruck Airport.

As an illustration of the tree-based wind direction forecasts at Innsbruck
Airport, a distributional tree of depth three is depicted in

Figure~\ref{fig:tree} depicts the resulting distributional tree, including
both the empirical (gray) and fitted von Mises (red) distribution of wind direction
in each terminal node. Based on the fitted location parameters $\hat \mu$, the subgroups
can be distinguished into the following wind regimes:
(1)~Up-valley winds blowing from the valley mouth towards the upper valley
(from east to west, nodes 4 and 5). (2) Downslope winds blowing across
the Alpine crest along the Wipp Valley towards Innsbruck (from south-east to
north-west, nodes 7 and 8). (3) Down-valley winds blowing in the direction
of the valley mouth (from west to east, nodes 12, 14, and 15). Node~11
captures observations with rather low wind speeds that cannot be distinguished
clearly into wind regimes and consequently are associated with a very low
estimated concentration $\hat \kappa$. In terms of covariates, the lagged
wind direction (``persistence'') is mostly responsible for distinguishing
the broad wind regimes listed above while the pressure gradients and wind
speed separate between subgroups with high vs.\ low precision.

@@

Distributional trees for circular responses are established by coupling
model-based recursive partitioning with the von Mises distribution.
The resulting trees can capture nonlinear changes, shifts, and potential interactions
in covariates without prespecification of such effects. This is modeling
wind direction in mountainous terrain where wind shifts can occur due to turns
a turn of the pressure gradient along a valley.

A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model
(see Schlosser et.~al, 2018 for a general discussion of distributional forests).

Another possible improvement for obtaining more parsimonious and more stable trees is to
consider splitting circular covariates into two circle segments by searching two
split points simultaneously (rather than sequentially at different depths as
for variable ``dd'' in Figure~\ref{fig:tree}).

\textsf{R} packages implementing the proposed methods are under development at
\texttt{https://R-Forge.R-project.org/projects/partykit/}.

@@

\item[Jammalamadaka, S. R., and Sengupta, A.] (2001).
     Topics in Circular Statistics.
     {\it World Scientific}.

@@

\item[Schlosser, L., Hothorn, T., and Zeileis, A.] (2018).
     Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain.
     arXiv:1804.02921, {\it arXiv.org E-Print Archive}.
